---
title: "History of football analysis"
author: "Roberto Galindo Ramírez"
output:
  html_document:
    df_print: paged
---

In a [recent analyisis](https://rgalindor.github.io/football-international/football-analysisi.html) we reviewed the performance of every National Team in the world. However evolution of the style of game is a fact that was not contemplated. So, in this post we can visualize how in FIFA World Cup every nation is evolving at different rates.

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(lubridate)
```

## Dataset

The first thing is to read files. I downloaded [this project](https://www.kaggle.com/martj42/international-football-results-from-1872-to-2017) at 2021-07-22 from [Kaggle](https://www.kaggle.com). 


```{r}
results <- read.csv("results.csv", encoding = "UTF-8")
```

This dataset contains data about $42k+$ football matches in the history of international encounters between national teams. So, let's take a little taste of the data:


```{r}
head(results)
```

## Analysis

One interesting thing is to take a look at the context of the matches, some of them could be not relevant at all, however, there is also **World cup** matches, continental tournaments, and so on:

```{r}
levels(as.factor(results$tournament)) -> tournaments
sample(tournaments,20)
```

Filtering by tournaments with at least 100 matches played in the history:

```{r}
results %>%
  group_by(tournament) %>%
  summarise(count=n()) %>%
  filter(count > 100) %>%
  select(tournament) -> popularCups

results %>%
  filter(tournament %in% popularCups$tournament) %>%
  ggplot(aes(x=tournament, fill=tournament)) +
  geom_bar() +
  coord_flip() +
  labs(title="Matches in tournaments") -> p 
ggplotly(p)
```


Now we need to process a little bit of the data to assign a standard way to provide points based on the outcome of every match:

| Points | Outcome |
|--|--|
| $3$ | Victory |
| $1$ | Tie |
| $0$ | Defeat |

> In FIFA scores, 2 points can be achieved by winning a shootout after a tied match, however, I ignored that for the following analysis

Let's take a look on how it looks now:

```{r}
results %>%
  mutate(tied=ifelse(home_score == away_score,TRUE,FALSE)) %>%
  mutate(home_points=ifelse(tied == TRUE,1,ifelse(home_score > away_score,3,0))) %>%
  mutate(away_points=ifelse(tied == TRUE,1,ifelse(home_score > away_score,0,3))) -> results

results %>%
  filter(grepl("FIFA World Cup",tournament)) -> worldCupResults
head(worldCupResults)
```

After this step we also need to transform a little bit the structure of this dataset in order to measure the performance of each National Team in this way:

 - **Performance** The points achieved by the team standardized by the total number of matches played.
 - **Offensive** The average number of goals scored by a team in each match played.
 - **Defense** The average number of goals received by a team in each match payed.

Then we can see how it looks (for tournaments that contain `"FIFA World Cup"` in its name).

```{r}
results %>%
  pivot_longer(c(home_team,away_team),names_to = "homeaway", values_to = "team") %>%
  mutate(points=ifelse(grepl("home",homeaway),home_points,away_points),
         goals=ifelse(grepl("home",homeaway),home_score,away_score),
         receivedGoals=ifelse(grepl("home",homeaway),away_score,home_score)) %>%
  select(date,tournament,country,team,points,goals,receivedGoals) -> results

results %>%
  filter(grepl("FIFA World Cup",tournament)) -> worldCupResults
```

### FIFA World Cup (and qualifiers)

The most interesting matches occur at [FIFA World Cup](https://www.fifa.com/tournaments/mens/worldcup). So we can focus on what happens in this tournament:

```{r}
worldCupResults %>% filter(!grepl("qualifi",tournament)) %>% mutate(yr=year(date)) %>% group_by(yr,team) %>% summarise( p=sum(points), goals=sum(goals), against=sum(receivedGoals),matches=n()) %>% mutate( performance=p/matches, ofensive=goals/matches, defense=against/matches) %>% ggplot(aes(x=yr, y=performance, fill=team)) + geom_bar(stat="identity") -> p
ggplotly(p)
```

Germany emerges as the best in performance over all the matches related to the **World Cup**. Is not a surprise at all, remember all of the _"goleadas"_ that has produced, in the qualifiers as well as in the knock-out matches in the final stages of the tournament.

Now we can take a look at what happens if we focus only on the final stage, I mean filtering out the qualifiers:

```{r}
worldCupResults %>% filter(!grepl("qualifi",tournament)) %>% mutate(yr=year(date)) %>% group_by(yr,team) %>% summarise( p=sum(points), goals=sum(goals), against=sum(receivedGoals),matches=n()) %>% mutate( performance=p/matches, ofensive=goals/matches, defense=against/matches) %>% filter(team %in% c("Mexico","Brazil","Argentina","Germany","France")) %>% ggplot(aes(x=yr, y=performance, color=team)) + geom_line() -> p
ggplotly(p)
```

```{r}
worldCupResults %>% filter(!grepl("qualifi",tournament)) %>% mutate(yr=year(date)) %>% group_by(yr,team) %>% summarise( p=sum(points), goals=sum(goals), against=sum(receivedGoals),matches=n()) %>% mutate( performance=p/matches, ofensive=goals/matches, defense=against/matches) %>% mutate(differenceGoal=ofensive-defense) %>% ggplot(aes(x=yr, color=team, y=differenceGoal)) + geom_line() -> p
ggplotly(p)
```






